Humor generation is one of the clearest examples of a language task that matters precisely because it is hard to verify automatically. A joke may be grammatically correct, semantically coherent, and even topically relevant while still failing at the central target: making a reader laugh. This makes humor generation a useful stress test for modern language-model post-training. In domains such as mathematics, coding, or formal logic, reinforcement learning can rely on explicit verification signals. In humor, by contrast, the output space is open-ended, the number of acceptable answers is large, and human judgments are noisy. The same prompt can admit many good jokes, while a dataset usually records only one or a few. These properties make humor an informative case for studying reasoning in non-verifiable domains.

The relevance of the topic is determined by three converging trends. First, recent surveys show that computational humor is moving from isolated rule-based or template-based systems toward large language model pipelines, yet the gap between fluent text and genuinely funny output remains substantial \citep{Loakman2025}. Second, post-training research increasingly suggests that supervised fine-tuning (SFT) and reinforcement learning optimize different capabilities: SFT is effective at imitation, whereas reinforcement learning is more promising when generalization beyond the training surface is required \citep{Chu2025}. Third, reasoning-focused reinforcement learning has recently produced strong results in domains with explicit verifiers, motivating the question of whether similar methods can be adapted to creative tasks where direct verification is impossible \citep{DeepSeek2025,Zhou2025,Yu2025}.

This thesis studies humor generation as a non-verifiable reasoning problem. The central idea is that humor does not only require a good final answer; it often benefits from exploratory reasoning over premises, associations, contrasts, and punchline formats. Multi-step humor systems already exploit this fact at inference time by explicitly decomposing the creative process into intermediate stages \citep{Tikhonov2024,ZhangJ2025,Kim2025}. However, such systems are architecturally restrictive: the number of stages, the reasoning style, and the intermediate control signals are designed manually. The alternative investigated in this thesis is to construct a training setting in which a model can learn how to reason for humor generation without requiring a symbolic verifier.

The object of the study is the post-training of large language models for creative text generation. The subject of the study is the design of a multi-reference verifier-free reinforcement learning objective for humor generation together with the supporting data pipeline needed to make such an objective operational.

The goal of the thesis is to develop a preliminary methodological and software foundation for training humor-generation models with verifier-free reinforcement learning, while remaining faithful to the actual implementation status of the project. In practical terms, the work aims to transform a large joke corpus into prompt-reference sets suitable for multi-reference training and evaluation, formulate the corresponding reinforcement learning objective, and analyze whether the constructed data artifacts are adequate for a later MRVF training stage.

To achieve this goal, the following tasks are addressed:
\begin{enumerate}
    \item analyze recent literature on computational humor, post-training, and verifier-free reasoning methods;
    \item define the research gap between current humor pipelines and reinforcement learning for non-verifiable domains;
    \item build a corpus-processing pipeline that converts raw jokes into weak prompts with multiple retrieved references;
    \item implement candidate-generation and evaluation components for later benchmark comparisons;
    \item formulate a multi-reference extension of verifier-free reinforcement learning and discuss its statistical consequences under incomplete references;
    \item evaluate the current pilot data pipeline using descriptive statistics, qualitative examples, and implementation analysis;
    \item outline the next experimental stage needed to turn the pilot into a full thesis-level training study.
\end{enumerate}

The working hypothesis of the thesis is that humor generation can be made more amenable to reasoning-oriented post-training if the learning problem is reframed from ``predict the single golden joke'' to ``assign probability mass to a set of acceptable jokes conditioned on a weak prompt.'' Under this hypothesis, the main obstacle is not only reward design, but also dataset structure: a useful training signal requires prompts that admit multiple valid references rather than isolated one-to-one joke mappings.

The scientific novelty of the work is threefold. First, it adapts the verifier-free reinforcement learning perspective of \citet{Zhou2025} and \citet{Yu2025} to a creative task where multiple valid outputs are expected by definition. Second, it makes explicit the difference between the original conceptual report and the current implemented system: rather than a full generative prompt inversion module, the present codebase realizes a retrieval-based approximation built on keywords, embedding search, and reference-set construction. Third, it provides an empirical pilot analysis of this approximation on real data, which is necessary before a more expensive reinforcement-learning stage can be justified.

The practical value of the work lies in the implemented software pipeline. The repository currently contains modular Python components for corpus collection, embedding generation, keyword extraction, reference retrieval, candidate generation, and pairwise evaluation. These components produce reusable parquet datasets and isolate stages that can later be reused for baseline generation, preference comparison, and reinforcement-learning experiments.

The empirical basis of the work consists of the combined joke corpus built from the Short Jokes and rJokes collections. In the current local state of the repository, the complete raw corpus contains 664{,}048 jokes, while the downstream pilot pipeline processes a smaller subset for prompt construction and reference analysis. This distinction is important: the thesis is preliminary not only in conceptual scope, but also in the current scale of the implemented experiments.

Methodologically, the thesis combines literature review, software engineering, corpus construction, embedding-based retrieval, mathematical derivation of an objective function, and descriptive analysis of intermediate artifacts. This combination reflects the actual status of the research problem: before large-scale reinforcement learning can be defended, the data representation, prompt formulation, and reference structure must first be stabilized.

The structure of the thesis follows the standard university requirements. Chapter 1 reviews the literature on humor generation, evaluation, and reasoning-oriented post-training. Chapter 2 presents the methodology, including the implemented pipeline and the formal MRVF objective. Chapter 3 reports the current pilot results, discusses strengths and limitations of the constructed dataset, and explains what remains to be implemented for the final thesis. The work concludes with a summary of findings, followed by the list of references and appendices.
