\section{Research scope and implementation status}

The methodology of this preliminary thesis has two layers. The first layer is an implemented data and evaluation pipeline, already present in the repository. The second layer is a proposed training objective, namely multi-reference verifier-free (MRVF) reinforcement learning, which is formulated mathematically but not yet fully implemented in code. This distinction is important for the scope of the present work. The repository already contains pipelines for corpus construction, embedding generation, keyword extraction, reference retrieval, candidate generation, and automatic pairwise evaluation. However, the training stage itself remains future work. Consequently, this chapter distinguishes between what is currently operational in the codebase and what is proposed as the next stage of the project.

At a high level, the implemented system constructs a weakly supervised multi-reference dataset from raw joke datasets. Jokes are first merged into a unified corpus, embedded with a dense retrieval model, and reduced to a small set of keywords with an emb. These keywords are then expanded into n-gram prompt groups, which are then used to retrieve semantically related jokes from the embedding index. The resulting references supports two immediate uses. First, it provides data for baseline generation and automatic evaluation. Second, it defines training and validation sets required by the proposed MRVF training objective.

\section{Dataset construction}

% TODO: Add dataset sizes here
The empirical basis is a merged corpus built from two public joke collections: the Short Jokes dataset and the rJokes dataset. In the repository, this stage is implemented in \texttt{src/pipelines/jokes.py}. The pipeline downloads the raw files from pinned source revisions, converts them into parquet format, and preserves source metadata for each joke. For every row, the stored fields are \texttt{id}, \texttt{text}, \texttt{source\_name}, \texttt{source\_filename}, and \texttt{source\_id}. After preprocessing the two corpora separately, they are concatenated and assigned a new global integer identifier.

This design is intentionally conservative. The pipeline does not perform aggressive text normalization beyond removing empty entries and standardizing storage format. For humor, such caution is desirable, because seemingly minor textual details, such as punctuation, wording, or line breaks may contribute to comedic effect. Retaining source metadata is also useful for later analysis of domain differences, duplicate detection, and possible source-specific biases.

Formally, let the merged corpus be denoted
\[
    \mathcal{J}=\{(i, y_i, s_i)\}_{i=1}^{N},
\]
where \(y_i\) is the joke text, \(i\) is a global identifier, and \(s_i\) collects source metadata. This corpus is the common starting point for all later stages.

\section{Dense embedding index}

The next stage embeds each joke in a dense semantic vector space. This is implemented in \texttt{src/pipelines/embeddings.py}. The current default configuration uses the model \texttt{Qwen3-Embedding-8B} with embedding dimensionality 1024, reduced from the original size of 4096 with Matryoshka Representation Learning. Embeddings are computed asynchronously in batches, written to parquet shards, and stored as pairs \((i, e_i)\), where \(i\) is the global joke identifier and \(e_i \in \mathbb{R}^{1024}\) is the corresponding embedding.

The role of this stage is twofold. First, the joke embeddings are later used to score candidate keywords by semantic relevance. Second, they define the vector index used for retrieving multi-reference joke sets. In other words, the embedding space is the shared representation on which both keyword extraction and reference retrieval depend.

\section{Keyword extraction as weak prompt retrieval}

Each joke is assumed to have a set of weak prompts that it plausibly answers. For example, ``write a joke about a chicken'' is considered a viable prompt for ``why did the chicken crossed the road'' joke. Keyword extraction stage is aimed at constructing weak prompts based on the keywords of jokes. This stage is implemented in \texttt{src/pipelines/keywords.py}.

For each joke \(y_i\), the pipeline first generates a candidate pool of n-grams with a \texttt{CountVectorizer} analyzer. Under the default configuration, this analyzer uses unigrams only, removes English stopwords, and keeps at most 64 candidate terms per joke. Let the resulting candidate set be
\[
    C(y_i)=\{c_{i1},\dots,c_{im}\}.
\]
Each candidate term is then embedded with the same embedding model used for jokes. If \(e(y_i)\) denotes the joke embedding and \(e(c)\) denotes the embedding of a candidate keyword, the relevance score is computed by cosine similarity:
\[
    s(c,y_i)=\frac{e(c)^\top e(y_i)}{\lVert e(c)\rVert\,\lVert e(y_i)\rVert}.
\]

Selecting only the most similar candidates would often yield redundant keywords. To reduce redundancy, the pipeline applies maximal marginal relevance (MMR), choosing multiple keywords while balancing relevance against diversity. Under the default configuration, the diversity coefficient is 0.7 and the number of retained keywords is \(3\). The output of this stage is therefore a compact keyword representation
\[
    K(y_i)=\{k_{i1},k_{i2},k_{i3}\},
\]
possibly with fewer than three items when the candidate pool is small.

Methodologically, these keywords should not be interpreted as full latent prompts. They are weaker objects, n-grams that capture topical or lexical anchors of the joke. This is a pragmatic approximation. It does not solve the joke-to-prompt inversion in a complete sense, but it yields a tractable representation from which multiple prompt options can be constructed.

\section{Reference construction by keyword grouping and retrieval}

\subsection{Keyword grouping and prompt rendering}

The reference construction stage is implemented in \texttt{src/pipelines/references.py}. For each joke keyword set \(K(y_i)\), the pipeline enumerates all unique keyword combinations whose size lies between \texttt{min\_keywords} and \texttt{max\_keywords}. Under the default configuration, these values are 1 and 2, so the pipeline constructs all one-keyword and two-keyword groups. If
\[
    K(y_i)=\{k_1,k_2,k_3\},
\]
then the generated groups are
\[
    \{k_1\},\ \{k_2\},\ \{k_3\},\ \{k_1,k_2\},\ \{k_1,k_3\},\ \{k_2,k_3\}.
\]
Each group is rendered into a natural language prompt using the template
\begin{quote}
    \ttfamily
    Write a joke using the following keyword(s): \{keywords\}
\end{quote}
This prompt serves two purposes: it is the textual form later given to generation models, and it is also the object embedded as a query for reference retrieval.

This design creates a controlled but diverse spectrum of prompts. Single keyword prompts are broad, topical and often admit many possible jokes. Two keyword prompts are more constrained and tend to retrieve more precise candidates. The present implementation stops at two keywords because larger groups would quickly become too specific, reducing the number of plausible distinct references per prompt.

\subsection{Reference retrieval with embedding index}

Once prompt strings are formed, the pipeline embeds them as queries and retrieves nearby jokes from a FAISS index built over the full joke-embedding collection. The retrieval index is an inverted file index (IVF). Before indexing and search, vectors are normalized, so that inner product corresponds to cosine similarity. Under the default configuration, the main parameters are:
\begin{itemize}
    \item embedding model: \texttt{Qwen3-Embedding-8B};
    \item dimensionality: 1024;
    \item \texttt{faiss\_nlist}: 4096;
    \item \texttt{faiss\_nprobe}: 32;
    \item \texttt{top\_k}: 5;
    \item \texttt{oversample}: 10;
    \item \texttt{min\_similarity}: 0.5.
\end{itemize}

For a rendered prompt \(x\), let \(q(x)\) be its normalized embedding and let \(e_j\) be the normalized embedding of joke \(y_j\). The retrieval score is then
\[
    s(x,y_j)=q(x)^\top e_j.
\]
The search procedure retrieves more than \(k\) candidates, filters out those whose similarity falls below the threshold, and retains the top \(k\) remaining items. The corresponding texts are then used as the observed reference set:
\[
    Y^{*}(x)=\{y_1,\dots,y_n\}, \qquad n \le 5.
\]

This stage implements the central point of the thesis: a prompt associated with multiple candidate joke references rather than with a single ground truth. Importantly, the retrieved set is not an exhaustively annotated set of all good jokes for \(x\). It is a retrieved approximation of that set. This matters both empirically and theoretically. Empirically, retrieved references may contain noise or omit valid alternatives. Theoretically, the incompleteness of \(Y^{*}(x)\) becomes a core part of the later MRVF analysis rather than a secondary inconvenience.

\subsection{Deduplication and splitting}

After retrieval, the resulting dataset contains rows of the form
\[
    (x_i, Y^{*}(x_i), S(x_i)),
\]
where \(x_i\) is the keyword-based prompt, \(Y^{*}(x_i)\) is the retrieved reference set, and \(S(x_i)\) is the vector of retrieval scores. Different source jokes can generate identical keyword groups, so the repository deduplicates the dataset by the \texttt{keywords} field, keeps the first occurrence, and reassigns row identifiers.

The deduplicated dataset is then split into train, validation, and test partitions. The current implementation uses a two-stage random split with \texttt{test\_fraction}=0.1, \texttt{validation\_fraction}=0.1, and a fixed random seed. Separate parquet shards are written for the full dataset and for each split.

\section{Candidate generation}

This stage supports candidate generation over the reference dataset and is implemented in \texttt{src/pipelines/candidates.py}. For a selected model and dataset split, the pipeline iterates over the prompt rows and generates one candidate joke using an OpenAI-compatible Chat Completion API. The generation prompt is the same prompt template used in reference construction:
\begin{quote}
    \ttfamily
    Write a joke using the following keyword(s): \{keywords\}
\end{quote}
Under the default configuration, candidate generation uses temperature \(1.0\) and allows up to 128 completion tokens. It is important keep the jokes short to simplify evaluation for both LLM-as-a-judge and human settings.

The output schema contains four fields: row identifier, keyword group, model name, and generated text. Candidate files are stored under a per-model, per-split directory structure. This organization is important for later evaluation, because it allows outputs from several models to be aligned by the same underlying identifier and prompt.

Methodologically, this candidate stage should be understood as a benchmarking layer. It provides the artifacts needed to compare baseline models (pre-trained base, instruct and thinking versions) and future trained checkpoints.

\section{Candidate evaluation}

\subsection{Pairwise comparison}

Automatic evaluation is implemented in \texttt{src/pipelines/evaluation.py}. The pipeline first concatenates candidate datasets from multiple models. It then groups candidates by reference identifier and model name, and constructs pairwise comparisons across all model pairs that produced outputs for the same prompt. Left-right ordering is randomized with a fixed seed in order to reduce position bias.

The evaluation prompt is deliberately simple, it contains the shared joke prompt together with a left and a right candidate. The system instruction tells the annotator to act as an judge of humor quality, choose which joke is funnier for a general audience, avoid ties, and return strict JSON. Under the default configuration, the judge model runs at temperature \(0\).

This evaluation layer captures only comparative funniness under a shared prompt. It does not yet implement the multidimensional evaluation discussed in the literature review, such as separate rubric-based judgments for prompt adherence, novelty, human-likeness, and funniness, nor does it yet include human annotation. These stages are future work as they depend on the complete training procedure.

\subsection{Leaderboard aggregation}

Pairwise judgments are aggregated into a leaderboard with wins, losses, number of comparisons, win rate, and Bradley-Terry scores. Let \(\mathcal{M}\) denote the set of evaluated models. For every judged pair \((m_a,m_b)\), the evaluator records a binary win for one side. The pipeline then estimates Bradley-Terry strengths and reports their logarithms as \texttt{bt\_score} values.

This aggregation is a sensible choice for humor because relative judgments are often more stable than absolute scores. Asking a judge to decide which of two jokes is funnier is usually easier than asking for a well-calibrated numerical rating. At the same time, this procedure inherits the limitations of the judge model. Since humor judgments by language models can diverge from human judgments, the current leaderboard should be interpreted as a fast and cheap proxy used for intermediate validation rather than as a definitive measure of humor quality.

\section{Theoretical framework for MRVF training}

The implemented pipeline above constructs prompt-conditioned reference sets, the present section specifies how such sets would enter a reasoning-oriented reinforcement-learning objective. The central claim is that humor generation is not well matched to the single-reference assumptions used in standard verifier-based or verifier-free reasoning RL. A weak prompt may admit many acceptable jokes, and the repository already makes this concrete by retrieving several references for the same keyword-based prompt. The theory below formalizes how to optimize against such set-valued supervision.

\subsection{Notation}

Let \(\mathcal{X}\), \(\mathcal{Z}\), and \(\mathcal{Y}\) denote the spaces of prompts, reasoning traces, and final answers, respectively. For a prompt \(x\in\mathcal{X}\),
\begin{itemize}
    \item \(z\in\mathcal{Z}\) denotes a reasoning trace,
    \item \(y\in\mathcal{Y}\) denotes the final generated joke,
    \item \(y^{*}\in\mathcal{Y}\) denotes a single reference answer,
    \item \(Y^{*}(x)\subset\mathcal{Y}\) denotes the \emph{observed} set of acceptable references associated with prompt \(x\),
    \item \(\pi_{\theta}\) denotes the policy, that is, the conditional probability model over text parameterized by \(\theta\).
\end{itemize}

More explicitly,
\[
    \pi_{\theta}(z,y\mid x)=P_{\theta}(Z=z,Y=y\mid X=x).
\]
The policy is factorized as
\[
    \pi_{\theta}(z,y\mid x)=\pi_{\theta}(z\mid x)\,\pi_{\theta}(y\mid x,z).
\]
When the model emits explicit reasoning tokens, \(z\) can be identified with the text inside a designated \texttt{<think>} segment and \(y\) with the final answer that follows it. When reasoning is not explicitly exposed in the output stream, the same factorization can still be read as a conceptual decomposition into a sampled intermediate trajectory \(z\) and a final answer conditioned on that trajectory. The advantage of this notation is that it separates two roles of the model: selecting a reasoning path and assigning probability mass to answers given that path.

For notational economy, the optimization objective below is written in terms of the observed set \(Y^{*}(x)\). When discussing incompleteness, it will be useful to imagine a larger latent set \(Y_{\mathrm{true}}(x)\supseteq Y^{*}(x)\) containing all genuinely acceptable jokes for prompt \(x\), even though that full set is never observed in practice.

\subsection{RL with verifiable rewards}

Reinforcement learning optimizes a policy by maximizing expected reward. Given an objective \(J(\theta)\), training proceeds by stochastic gradient ascent or, equivalently, by gradient descent on the negative objective. The critical design choice is therefore the reward function: it specifies which generations should receive positive signals.

In verifiable domains, reward can be assigned by an external procedure that does not depend on subjective human judgment. For code, a generated program can be executed against a test suite, for mathematics, a final answer can be parsed and compared against a canonical solution. In such cases the event ``the answer is correct'' has an meaning external to the language model itself. Recent reasoning models rely heavily on this structure for training \citep{DeepSeek2025,Lightman2023}.

Assume first the standard single-reference setting: for each prompt \(x\), there is exactly one verifiable target \(y^{*}\). Define the reward
\[
    R(y,y^{*})=\mathbb{I}\{y=y^{*}\},
\]
where \(\mathbb{I}\{\cdot\}\) is the indicator function. The objective is then
\[
    J_{\mathrm{RLVR}}(\theta\mid x,y^{*})
    =
    \mathbb{E}_{z\sim\pi_{\theta}(z\mid x)}
    \mathbb{E}_{y\sim\pi_{\theta}(y\mid x,z)}
    \big[R(y,y^{*})\big].
\]
Equivalently, in expanded summation form,
\[
    J_{\mathrm{RLVR}}(\theta\mid x,y^{*})
    =
    \sum_{z\in\mathcal{Z}}\sum_{y\in\mathcal{Y}}
    \pi_{\theta}(z\mid x)\pi_{\theta}(y\mid x,z)\,R(y,y^{*}).
\]

To optimize this expectation one uses the score-function identity, also known as the REINFORCE identity:
\[
    \nabla_{\theta}\,\mathbb{E}_{u\sim p_{\theta}(u)}[f(u)]
    =
    \mathbb{E}_{u\sim p_{\theta}(u)}
    \big[f(u)\nabla_{\theta}\log p_{\theta}(u)\big].
\]
This identity follows directly from the relation \(\nabla_{\theta}p_{\theta}(u)=p_{\theta}(u)\nabla_{\theta}\log p_{\theta}(u)\). Applying it to the joint distribution of \((z,y)\) yields
\begin{align*}
    \nabla_{\theta}J_{\mathrm{RLVR}}
     & =
    \nabla_{\theta}
    \sum_{z,y}\pi_{\theta}(z,y\mid x)\,R(y,y^{*})                                             \\
     & =
    \sum_{z,y}R(y,y^{*})\,\nabla_{\theta}\pi_{\theta}(z,y\mid x)                              \\
     & =
    \sum_{z,y}\pi_{\theta}(z,y\mid x)\,R(y,y^{*})\,\nabla_{\theta}\log\pi_{\theta}(z,y\mid x) \\
     & =
    \mathbb{E}_{z,y}\Big[
        R(y,y^{*})\,\nabla_{\theta}\log\pi_{\theta}(z,y\mid x)
        \Big].
\end{align*}
Using the factorization of the joint policy,
\[
    \log\pi_{\theta}(z,y\mid x)
    =
    \log\pi_{\theta}(z\mid x)+\log\pi_{\theta}(y\mid x,z),
\]
hence
\[
    \nabla_{\theta}\log\pi_{\theta}(z,y\mid x)
    =
    \nabla_{\theta}\log\pi_{\theta}(z\mid x)
    +
    \nabla_{\theta}\log\pi_{\theta}(y\mid x,z).
\]
Therefore
\[
    \nabla_{\theta}J_{\mathrm{RLVR}}
    =
    \mathbb{E}_{z,y}\Big[
        R(y,y^{*})
        \big(
        \nabla_{\theta}\log\pi_{\theta}(z\mid x)
        +
        \nabla_{\theta}\log\pi_{\theta}(y\mid x,z)
        \big)
        \Big].
\]

This expression has a simple interpretation. A sampled pair \((z,y)\) receives reward only when the verifier declares \(y\) correct. When that occurs, the update increases both the probability of the reasoning trace that led to the answer and the probability of the answer itself under that trace

\subsection{VeriFree}

Humor does not provide such an external verifier, and more generally many open-ended generation tasks do not admit a deterministic correctness test. VeriFree addresses this gap by observing that in the single-reference case the expected exact-match reward can be rewritten as a conditional probability \citep{Zhou2025}.

Fix \(x\) and \(z\). Then
\begin{align*}
    \mathbb{E}_{y\sim\pi_{\theta}(\cdot\mid x,z)}
    \big[\mathbb{I}\{y=y^{*}\}\big]
     & =
    \sum_{y\in\mathcal{Y}}
    \pi_{\theta}(y\mid x,z)\,\mathbb{I}\{y=y^{*}\} \\
     & =
    \pi_{\theta}(y^{*}\mid x,z).
\end{align*}
Substituting this identity back into the RLVR objective gives
\[
    J_{\mathrm{RLVR}}(\theta\mid x,y^{*})
    =
    \mathbb{E}_{z\sim\pi_{\theta}(z\mid x)}
    \big[\pi_{\theta}(y^{*}\mid x,z)\big].
\]
This suggests defining the verifier-free reward
\[
    r_{\theta}(x,z,y^{*}) := \pi_{\theta}(y^{*}\mid x,z),
\]
and the corresponding objective
\[
    J_{\mathrm{VF}}(\theta\mid x,y^{*})
    :=
    \mathbb{E}_{z\sim\pi_{\theta}(z\mid x)}
    \big[r_{\theta}(x,z,y^{*})\big].
\]
Under the single-reference assumption,
\[
    J_{\mathrm{VF}}(\theta\mid x,y^{*}) \equiv J_{\mathrm{RLVR}}(\theta\mid x,y^{*})
\]
as scalar objectives. The importance of this reformulation is practical as well as conceptual: no explicit decoding of \(y\) is required for reward computation, and \(\pi_{\theta}(y^{*}\mid x,z)\) can be computed by teacher forcing on the reference sequence, which is conveniently parallelizable.

To obtain the policy gradient, write
\[
    J_{\mathrm{VF}}(\theta\mid x,y^{*})
    =
    \sum_{z\in\mathcal{Z}} \pi_{\theta}(z\mid x)\,r_{\theta}(x,z,y^{*}).
\]
Differentiating by the product rule,
\begin{align*}
    \nabla_{\theta}J_{\mathrm{VF}}
     & =
    \sum_{z}
    \Big[
        \nabla_{\theta}\pi_{\theta}(z\mid x)\,r_{\theta}(x,z,y^{*})
        +
        \pi_{\theta}(z\mid x)\,\nabla_{\theta}r_{\theta}(x,z,y^{*})
    \Big] \\
     & =
    \sum_{z}\pi_{\theta}(z\mid x)
    \Big[
        r_{\theta}(x,z,y^{*})\,\nabla_{\theta}\log\pi_{\theta}(z\mid x)
        +
        \nabla_{\theta}r_{\theta}(x,z,y^{*})
        \Big].
\end{align*}
Since \(r_{\theta}(x,z,y^{*})=\pi_{\theta}(y^{*}\mid x,z)\),
\[
    \nabla_{\theta}r_{\theta}(x,z,y^{*})
    =
    \pi_{\theta}(y^{*}\mid x,z)\,\nabla_{\theta}\log\pi_{\theta}(y^{*}\mid x,z)
    =
    r_{\theta}(x,z,y^{*})\,\nabla_{\theta}\log\pi_{\theta}(y^{*}\mid x,z).
\]
Hence
\[
    \nabla_{\theta}J_{\mathrm{VF}}
    =
    \mathbb{E}_{z\sim\pi_{\theta}(z\mid x)}\Big[
        r_{\theta}(x,z,y^{*})
        \big(
        \nabla_{\theta}\log\pi_{\theta}(z\mid x)
        +
        \nabla_{\theta}\log\pi_{\theta}(y^{*}\mid x,z)
        \big)
        \Big].
\]

This decomposition is informative. The first term,
\[
    \mathbb{E}_{z}\big[r_{\theta}(x,z,y^{*})\,\nabla_{\theta}\log\pi_{\theta}(z\mid x)\big],
\]
is a score-function term over sampled reasoning traces. It assigns higher probability to traces under which the model itself places higher conditional mass on the reference answer. The second term,
\[
    \mathbb{E}_{z}\big[\nabla_{\theta}r_{\theta}(x,z,y^{*})\big],
\]
is a direct teacher-forced gradient on the reference sequence. Relative to RLVR, VeriFree removes noise from sampling the final answer \(y\): the estimator still samples \(z\), but not the answer used for reward evaluation. This is precisely why VeriFree is attractive in domains where a verifier is unavailable but a single reference answer exists.

\subsection{Multi-reference VeriFree}

The single-reference assumption is inappropriate for humor. A prompt such as ``write a joke using the keywords cat, dog'' does not have one correct completion, it has many plausible completions differing in setup, punchline mechanism, tone, and level of absurdity. In fact, the implemented repository is explicitly designed to construct several references for the same weak prompt. The reward must therefore be defined on a set of acceptable answers rather than on one.

Let \(Y^{*}(x)\subset\mathcal{Y}\) denote the observed set of acceptable references for prompt \(x\). Define the multi-reference verified reward
\[
    R(y,Y^{*})=\mathbb{I}\{y\in Y^{*}\}.
\]
The corresponding set-valued exact-match objective is
\[
    J_{\mathrm{MR}}(\theta\mid x,Y^{*})
    =
    \mathbb{E}_{z\sim\pi_{\theta}(z\mid x)}
    \mathbb{E}_{y\sim\pi_{\theta}(\cdot\mid x,z)}
    \big[\mathbb{I}\{y\in Y^{*}\}\big].
\]
For fixed \(x\) and \(z\),
\begin{align*}
    \mathbb{E}_{y\sim\pi_{\theta}(\cdot\mid x,z)}
    \big[\mathbb{I}\{y\in Y^{*}\}\big]
     & =
    \sum_{y\in\mathcal{Y}}
    \pi_{\theta}(y\mid x,z)\,\mathbb{I}\{y\in Y^{*}\} \\
     & =
    \sum_{y\in Y^{*}}
    \pi_{\theta}(y\mid x,z).
\end{align*}
This motivates the multi-reference verifier-free reward
\[
    r_{\theta}(x,z,Y^{*})
    :=
    \sum_{y\in Y^{*}}
    \pi_{\theta}(y\mid x,z).
\]
The quantity \(r_{\theta}(x,z,Y^{*})\) lies in \([0,1]\): it is the total probability mass that the policy assigns, under reasoning trace \(z\), to the observed reference set. Equivalently, it is the probability of the event that the generated answer falls inside that set. The resulting objective is
\[
    J_{\mathrm{MRVF}}(\theta\mid x,Y^{*})
    =
    \mathbb{E}_{z\sim\pi_{\theta}(z\mid x)}
    \big[r_{\theta}(x,z,Y^{*})\big].
\]

At the scalar level, MRVF plays the same role for a set-valued reward that VeriFree plays for a single exact-match reward. However, the gradient structure is not merely the single-reference formula with \(y^{*}\) replaced by a set symbol. The answer-side derivative changes in an essential way.

Indeed,
\[
    J_{\mathrm{MRVF}}(\theta\mid x,Y^{*})
    =
    \sum_{z}\pi_{\theta}(z\mid x)\,r_{\theta}(x,z,Y^{*}),
\]
so
\begin{align*}
    \nabla_{\theta}J_{\mathrm{MRVF}}
     & =
    \sum_{z}
    \Big[
        \nabla_{\theta}\pi_{\theta}(z\mid x)\,r_{\theta}(x,z,Y^{*})
        +
        \pi_{\theta}(z\mid x)\,\nabla_{\theta}r_{\theta}(x,z,Y^{*})
    \Big] \\
     & =
    \mathbb{E}_{z\sim\pi_{\theta}(z\mid x)}
    \Big[
        r_{\theta}(x,z,Y^{*})\,\nabla_{\theta}\log\pi_{\theta}(z\mid x)
        +
        \nabla_{\theta}r_{\theta}(x,z,Y^{*})
        \Big].
\end{align*}
Now expand the reference-side derivative:
\begin{align*}
    \nabla_{\theta}r_{\theta}(x,z,Y^{*})
     & =
    \nabla_{\theta}
    \sum_{y\in Y^{*}}\pi_{\theta}(y\mid x,z) \\
     & =
    \sum_{y\in Y^{*}}
    \nabla_{\theta}\pi_{\theta}(y\mid x,z)   \\
     & =
    \sum_{y\in Y^{*}}
    \pi_{\theta}(y\mid x,z)\,\nabla_{\theta}\log\pi_{\theta}(y\mid x,z).
\end{align*}
Therefore,
\[
    \nabla_{\theta}J_{\mathrm{MRVF}}
    =
    \mathbb{E}_{z\sim\pi_{\theta}(z\mid x)}
    \Bigg[
        r_{\theta}(x,z,Y^{*})\,\nabla_{\theta}\log\pi_{\theta}(z\mid x)
        +
        \sum_{y\in Y^{*}}
        \pi_{\theta}(y\mid x,z)\,\nabla_{\theta}\log\pi_{\theta}(y\mid x,z)
        \Bigg].
\]

It is sometimes useful to rewrite the second term in normalized form. Whenever \(r_{\theta}(x,z,Y^{*})>0\), define
\[
    w_{\theta}(y\mid x,z,Y^{*})
    :=
    \frac{\pi_{\theta}(y\mid x,z)}{r_{\theta}(x,z,Y^{*})}
    \,\mathbb{I}\{y\in Y^{*}\}.
\]
Then \(w_{\theta}(\cdot\mid x,z,Y^{*})\) is a probability distribution over the observed reference set, and
\[
    \nabla_{\theta}r_{\theta}(x,z,Y^{*})
    =
    r_{\theta}(x,z,Y^{*})
    \sum_{y\in Y^{*}}
    w_{\theta}(y\mid x,z,Y^{*})
    \nabla_{\theta}\log\pi_{\theta}(y\mid x,z).
\]
Equivalently,
\[
    \nabla_{\theta}\log r_{\theta}(x,z,Y^{*})
    =
    \sum_{y\in Y^{*}}
    w_{\theta}(y\mid x,z,Y^{*})
    \nabla_{\theta}\log\pi_{\theta}(y\mid x,z),
\]
and hence
\[
    \nabla_{\theta}J_{\mathrm{MRVF}}
    =
    \mathbb{E}_{z\sim\pi_{\theta}(z\mid x)}
    \Big[
        r_{\theta}(x,z,Y^{*})
        \big(
        \nabla_{\theta}\log\pi_{\theta}(z\mid x)
        +
        \nabla_{\theta}\log r_{\theta}(x,z,Y^{*})
        \big)
        \Big].
\]

This form clarifies the difference from single-reference VeriFree. In VeriFree, the direct term is the gradient of \(\log\pi_{\theta}(y^{*}\mid x,z)\) for one reference. In MRVF, the direct term is the gradient of \(\log \sum_{y\in Y^{*}}\pi_{\theta}(y\mid x,z)\). The update therefore does not privilege one reference a priori, it increases total mass on the observed set, with each reference weighted by the current policy mass assigned to it. If one reference already dominates, the gradient concentrates there. If mass is spread across several references, the gradient reinforces them jointly. This is why MRVF is not simply a notational variant of VeriFree: the optimization geometry changes from a single target string to a log-sum over many targets.

\subsection{Incomplete observed references}

The observed set \(Y^{*}(x)\) is still not the full set of good jokes. Two distinct forms of incompleteness arise.

\paragraph{Structural incompleteness.}
Even if the repository retrieves several references for a prompt, there may exist many additional jokes in the latent set \(Y_{\mathrm{true}}(x)\) that are also valid but unobserved. This is unavoidable in creative domains. A finite dataset cannot enumerate all acceptable punchlines for a weak prompt.

\paragraph{Computational incompleteness.}
Even within the observed set \(Y^{*}(x)\), computing
\[
    r_{\theta}(x,z,Y^{*})=\sum_{y\in Y^{*}}\pi_{\theta}(y\mid x,z)
\]
may become expensive when \(|Y^{*}|\) is large. One may therefore evaluate the sum on a sampled subset \(Y\subseteq Y^{*}\).

Suppose \(Y\) is sampled uniformly without replacement from \(Y^{*}\), and each reference \(y\in Y^{*}\) has inclusion probability
\[
    \rho = P(y\in Y)=\frac{|Y|}{|Y^{*}|}.
\]
Define the subset reward
\[
    r_{\theta}(x,z,Y)
    :=
    \sum_{y\in Y}\pi_{\theta}(y\mid x,z)
    =
    \sum_{y\in Y^{*}}
    \mathbb{I}\{y\in Y\}\,\pi_{\theta}(y\mid x,z).
\]
Taking expectation over the random subset,
\begin{align*}
    \mathbb{E}_{Y}\big[r_{\theta}(x,z,Y)\big]
     & =
    \sum_{y\in Y^{*}}
    \mathbb{E}_{Y}\big[\mathbb{I}\{y\in Y\}\big]\,
    \pi_{\theta}(y\mid x,z)                  \\
     & =
    \sum_{y\in Y^{*}}
    P(y\in Y)\,\pi_{\theta}(y\mid x,z)       \\
     & =
    \rho
    \sum_{y\in Y^{*}}\pi_{\theta}(y\mid x,z) \\
     & =
    \rho\,r_{\theta}(x,z,Y^{*}).
\end{align*}
Thus the subset reward is proportional in expectation to the full observed-set reward.

This is where GRPO baselining becomes relevant. Suppose that for a fixed prompt \(x\) all sampled rewards in a group are multiplied by the same positive factor \(\rho_x\), for example because only a fraction of the observed references is used. If one standardizes rewards within the group,
\[
    \widetilde A_i
    =
    \frac{\hat r_i - \mu_x}{\sigma_x+\varepsilon},
    \qquad
    \mu_x=\frac{1}{G}\sum_{j=1}^{G}\hat r_j,
\]
then multiplying every \(\hat r_i\) by \(\rho_x>0\) multiplies both the numerator and denominator by the same factor, leaving \(\widetilde A_i\) unchanged. This is the sense in which GRPO group standardization is scale-invariant. It mitigates reward-magnitude shifts in the score-function term, although it does not recover information lost through missing references and does not by itself solve the bias of the direct reference gradient.

Another fundamental issue is the false-negative problem. Optimizing
\[
    r_{\theta}(x,z,Y^{*})=\sum_{y\in Y^{*}}\pi_{\theta}(y\mid x,z)
\]
increases probability mass on observed references. Because the conditional distribution over all strings sums to one,
\[
    \sum_{y\in\mathcal{Y}}\pi_{\theta}(y\mid x,z)=1,
\]
that increased mass must come from \(\mathcal{Y}\setminus Y^{*}\). Unfortunately, this complement contains not only bad jokes but also unobserved good jokes from \(Y_{\mathrm{true}}(x)\setminus Y^{*}(x)\). In other words, the training signal can inadvertently suppress valid but unseen humorous continuations. This risk is not a defect of the gradient derivation, it is a structural consequence of learning from incomplete supervision in an intrinsically many-solution domain.

To reduce this risk, the proposed optimization includes a Kullback-Leibler penalty relative to a reference policy \(\pi_{\mathrm{ref}}\):
\[
    \max_{\theta}
    \;
    \mathbb{E}_{x,z}\big[r_{\theta}(x,z,Y^{*}(x))\big]
    -
    \beta\,
    \mathbb{E}_{x}\Big[
        \mathrm{KL}\big(
        \pi_{\theta}(\cdot\mid x)\,\|\,\pi_{\mathrm{ref}}(\cdot\mid x)
        \big)
        \Big].
\]
Equivalently, one may view the problem in constrained form,
\[
    \max_{\theta}\;
    \mathbb{E}_{x,z}\big[r_{\theta}(x,z,Y^{*}(x))\big]
    \quad
    \text{s.t.}
    \quad
    \mathbb{E}_{x}
    \Big[
        \mathrm{KL}\big(
        \pi_{\theta}(\cdot\mid x)\,\|\,\pi_{\mathrm{ref}}(\cdot\mid x)
        \big)
        \Big]
    \le \varepsilon.
\]
In the present context the KL term is not merely generic regularization. Its methodological role is to discourage overly aggressive redistribution of probability mass away from the reference model when the observed reference sets are known to be incomplete.

\subsection{Proposed training loop}

The intended training stage samples multiple reasoning traces for the same prompt and uses their rewards to update the reasoning policy. For a prompt \(x\), sample a group
\[
    z_1,\dots,z_G \sim \pi_{\theta}(z\mid x).
\]
For each \(z_i\), compute the multi-reference reward
\[
    \hat r_i
    :=
    r_{\theta}(x,z_i,Y^{*}(x))
    =
    \sum_{y\in Y^{*}(x)}
    \pi_{\theta}(y\mid x,z_i).
\]
In practice, each \(\pi_{\theta}(y\mid x,z_i)\) is obtained by teacher forcing on the reference sequence \(y\). Since these sequence probabilities are very small, implementation should accumulate token log-probabilities and, where necessary, use stabilization such as \(\log\sum\exp\) before converting back to probabilities or normalized weights.

The gradient naturally decomposes into three parts:
\[
    \nabla_{\theta}J_{\mathrm{MRVF}}
    =
    g_{z}+g_{r}+g_{\mathrm{KL}}.
\]

\paragraph{Reasoning term.}
The score-function part over \(z\) is estimated from sampled traces and therefore benefits from variance reduction. A clean choice is the leave-one-out baseline
\[
    b_i
    =
    \frac{1}{G-1}\sum_{j\neq i}\hat r_j,
    \qquad
    A_i=\hat r_i-b_i.
\]
The corresponding estimator is
\[
    g_{z}
    =
    \frac{1}{G}\sum_{i=1}^{G}
    A_i\,\nabla_{\theta}\log\pi_{\theta}(z_i\mid x).
\]
This is the same logic that motivates RLOO estimators: each trace is compared against the other traces sampled for the same prompt. If desired, one may further normalize the \(A_i\) values within the group to reduce prompt-specific reward-scale variation, in the spirit of GRPO. Such normalization improves robustness but is no longer strictly unbiased because the scale estimate itself depends on the sampled group.

\paragraph{Reference term.}
For the direct answer-side contribution, the correct observed-set gradient is
\[
    g_{r}
    =
    \frac{1}{G}\sum_{i=1}^{G}
    \nabla_{\theta}r_{\theta}(x,z_i,Y^{*}(x))
    =
    \frac{1}{G}\sum_{i=1}^{G}
    \sum_{y\in Y^{*}(x)}
    \pi_{\theta}(y\mid x,z_i)\,
    \nabla_{\theta}\log\pi_{\theta}(y\mid x,z_i).
\]
Equivalently, using the normalized within-set weights
\[
    w_{i,y}
    :=
    \frac{\pi_{\theta}(y\mid x,z_i)}
    {\hat r_i}
    \,\mathbb{I}\{y\in Y^{*}(x)\},
\]
one may write
\[
    g_{r}
    =
    \frac{1}{G}\sum_{i=1}^{G}
    \hat r_i
    \sum_{y\in Y^{*}(x)}
    w_{i,y}\,
    \nabla_{\theta}\log\pi_{\theta}(y\mid x,z_i),
\]
whenever \(\hat r_i>0\). This form makes the geometry explicit: each trace receives a total weight \(\hat r_i\), and within that trace the gradient is distributed across references according to the current policy mass on each reference.

\paragraph{KL term.}
Finally,
\[
    g_{\mathrm{KL}}
    =
    -\beta\,
    \nabla_{\theta}
    \mathrm{KL}\big(
    \pi_{\theta}(\cdot\mid x)\,\|\,\pi_{\mathrm{ref}}(\cdot\mid x)
    \big).
\]
The full update for a minibatch averages these prompt-level contributions over sampled prompts.

The practical importance of the implemented pipeline is now clear. Without prompt-conditioned multi-reference sets \(Y^{*}(x)\), neither \(\hat r_i\) nor \(g_r\) can be computed. The data pipeline is therefore not auxiliary to the theory, it provides the observed structure on which MRVF optimization depends.

\subsection{Note on baselining}

Variance reduction is straightforward only for the score-function term. The reason is that \(z\) is genuinely sampled from \(\pi_{\theta}(z\mid x)\). If \(b(x,z_{-i})\) is any baseline that does not depend on the sampled variable \(z_i\) itself, then
\begin{align*}
    \mathbb{E}_{z_i\sim\pi_{\theta}(\cdot\mid x)}
    \big[
        b(x,z_{-i})\,\nabla_{\theta}\log\pi_{\theta}(z_i\mid x)
        \big]
     & =
    b(x,z_{-i})
    \sum_{z_i}\pi_{\theta}(z_i\mid x)\,
    \nabla_{\theta}\log\pi_{\theta}(z_i\mid x)       \\
     & =
    b(x,z_{-i})
    \sum_{z_i}\nabla_{\theta}\pi_{\theta}(z_i\mid x) \\
     & =
    b(x,z_{-i})\,\nabla_{\theta}1                    \\
     & = 0.
\end{align*}
This is why the leave-one-out baseline is unbiased for the reasoning term.

The reference term is different. It is not an expectation over sampled answers \(y\sim\pi_{\theta}(\cdot\mid x,z)\), it is a direct derivative of the summed reference probability. Consequently there is no identity of the form
\[
    \mathbb{E}_{y}\big[\nabla_{\theta}\log\pi_{\theta}(y\mid x,z)\big]=0
\]
available for centering. For the single-reference case, for example,
\[
    \mathbb{E}_{z}
    \big[
        (r_{\theta}(x,z,y^{*})-b(x))
        \nabla_{\theta}\log\pi_{\theta}(y^{*}\mid x,z)
        \big]
\]
equals
\[
    \mathbb{E}_{z}
    \big[
        r_{\theta}(x,z,y^{*})
        \nabla_{\theta}\log\pi_{\theta}(y^{*}\mid x,z)
        \big]
    -
    b(x)\,
    \mathbb{E}_{z}
    \big[
        \nabla_{\theta}\log\pi_{\theta}(y^{*}\mid x,z)
        \big],
\]
and the second term is not generally zero. The same issue persists in the multi-reference case with \(\nabla_{\theta}\log r_{\theta}(x,z,Y^{*})\). Therefore baselines that are harmless for the score-function term become biased if naively applied to the direct reference gradient.

This distinction is important for the design of the training algorithm. The reasoning term should use variance-reduction machinery such as leave-one-out baselines or carefully chosen group normalization. The reference term should instead be computed as directly as possible from the observed-set probability, accepting that it is a deterministic but potentially biased proxy for the true reward produced by \(Y_{\mathrm{true}}(x)\).

Taken together, these observations explain the overall structure of the proposal. MRVF inherits the key computational advantage of verifier-free learning: it avoids external verifiers and explicit reward sampling over final answers but it must be adapted to the multi-reference, incomplete characteristic of humor. The resulting objective is mathematically tractable, but its limitations are part of the thesis contribution: incompleteness, false negatives, and estimator design are central to the problem.
